---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Return of the BMI (Data processing and statistical analysis in R)

<!-- badges: start -->
<!-- badges: end -->

This repository is an exercise in processing data and performing statistical analysis in R using a dataset from a Pakistan school nutrition survey of students older than 10 years old.

By this time, you should already know how to clone this repository into your local machine using RStudio. If you need a refresher, the following tutorial shows the steps on how to do this - https://oxford-ihtm.io/ihtm-handbook/clone-repository.html.

## Instructions for the assignment

The following tasks have been designed to help students get familiar with processing of data and performing statistical analysis in R.

The students are expected to go through the tasks and appropriately write R code/script to fulfil the tasks and/or to answer the question/s being asked within the tasks. R code/script should be written inside a single R file named `return_of_the_bmi.R` and saved in the project's root directory.

## Task 1: Honey, I lost the data!

For some reason, the data `school_nutrition.csv` has been lost and is not found within the `data` folder within the project directory.

Thankfully, this data is available via Oxford IHTM's GitHub repository called `teaching_datasets` found at https://github.com/OxfordIHTM/teaching_datasets. In this repository, there are two files in different formats that both contain the dataset we need for this exercise. There is the CSV file called `school_nutrition.csv` which is the exact same dataset that you had in the previous exercise and there is also the XLSX file called `school_nutrition.xlsx` which is the exact same dataset but as an Excel file.

For this task, you should look into ways to get back the school nutrition data and read it into R as we did previously.

1. Looking deeper into the `read.table()` function (`?read.table`), is there a way to read a CSV file directly from a URL or web link? If so, write code that does this.

2. You may want to be able to download the dataset and save it back into the `data` directory. The function `download.file()` (`?download.file`) can be used for this purpose. Write code that would download the CSV file into the `data` directory of the project and then read the CSV file from the `data` directory into R.

3. It would be good to learn how to read data from an XLSX file as it will not be uncommon these days to have data in XLSX format. For this purpose, the `openxlsx` package will be useful, specifically its `read.xlsx()` function. Install `openxlsx` (if you haven't done so already), load it into your environment (`library(openxlsx)`), and then read up on the `read.xlsx()` function (`?read.xlsx`) to figure out how to read the `student_nutrition.xlsx` file via URL using the `read.xlsx()` function. Write code that shows these steps.

## Task 2: Calculate BMI

Now that you have the school nutrition data into your R environment, write code that would calculate the BMI for each student in the dataset and then add this value as a new variable called `bmi` in the student nutrition dataset object.

To make this more challenging, please write a function that will specifically make this calculation of BMI. You have done this already earlier so make sure you pick up on feedback provided earlier to improve the function you have written.

## Task 3: Calculating age-standardised BMI

Unlike adults, BMI for children need to be age-standardised. For this, calculation of z-scores is the statistical approach used.

Read up on z-scores and how they are calculated. Then try to find an R package that has function/s that will calculate z-scores for BMI that adjusts for sex and age. Once you have found such package, learn how to use the specific function/s and then implement this in your code by calculating the BMI-for-age z-score for each student and then add this value as a new variable called `bfaz` in the student nutrition dataset object.

Once you have calculated `bfaz`, create an R function that will classify the BMI status of each student based on their BMI-for-age z-score. Following are the cut-offs for BMI-for-age z-score based on the WHO Growth Standards:

```{r bfaz-cutoff, echo = FALSE}
bmi <- c(
  "BMI-for-age z-score > 2 SD",
  "2 SD >= BMI-for-age z-score > 1 SD",
  "1 SD >= BMI-for-age z-score >= -2 SD",
  "-2 SD > BMI-for-age z-score >= -3 SD",
  "BMI-for-age z-score < -3 SD"
)

classification <- c(
  "Obese",
  "Overweight",
  "Normal",
  "Thin",
  "Severely thin"
)

data.frame(bmi, classification) |>
  knitr::kable(col.names = c("BMI-for-age z-score range", "Classification"))
```

## Task 4: Show and test the normality of the distribution of the BMI-for-age z-score variable

Write code that tests the normality of the BMI variable and answer the following questions:

* Is the BMI-for-age z-score of all students in the sample normally distributed?

* Is the BMI-for-age z-score of all the male students in the sample normally distributed?

* Is the BMI-for-age z-score of all the female students in the sample normally distributed?

Please write a script that shows normality **graphically** and through **formal statistical test/s**.

## Task 5: Summarising BMI data

Write code that summarises the data as follows:

* Mean value of BMI-for-age z-score for male and female students and overall;

* Number and proportion of children who are undernourished based on BMI-for-age z-score (at least thin);

* Number and proportion of children who are over-nourished based on BMI-for-age z-score (at least overweight).

Please show your results as rectangular data (data.frame or table format) with the rows for values for males, females, and overall and columns showing the summaries of interest (mean, sum, proportion) as shown in the dummy table below:


**grouping** | **mean_bfaz** | **n_undernourished** | **prop_undernourished** | **n_overnourished** | **prop_overnourished**
:--- | :--- | :--- | :--- | :---
**females** | &nbsp; | &nbsp; | &nbsp; | &nbsp;
**males** | &nbsp; | &nbsp; | &nbsp; | &nbsp;
**overall** | &nbsp; | &nbsp; | &nbsp; | &nbsp;

## Task 6: Statistical tests

Knowing what you know now about performing statistical tests from your Epi/Stats lectures, please answer the following questions by writing appropriate R code/script:

* Is there a difference between the mean BMI-for-age z-score of male students compared to the female students?

* Is there a difference between the proportion of undernourished male students compared to undernourished female students?

* Is there a difference between the proportion of overnourished male students compared to overnourished female students?








